Open-Set Domain Adaptation through Self-Supervision

### 2022-02-08 paper/main.tex
Some additions

>>>>

\section{Method}
\label{sec:method}

\subsection{Problem Formulation}
\label{sec:problemformulation}
Our starting point is the source dataset, defined as $\mathcal{D}_s = \{({\bf x}_i^s, y_i^s)\}_{i=1}^{N_S} \sim p_s$, where each element ${\bf x}_i^s$ belonging to any $y_i^s$ is a sample from the source domain $S$.
This dataset has a target counterpart, $\mathcal{D}_t = \{{\bf x}_i^t\}_{i=1}^{N_t} \sim p_t$ which is unlabeled.
In OSDA we have that $p_s \neq p_t$.
The source dataset $\mathcal{D}_s$ is associated with a set of known classes, $\mathcal{C}_s$, which can also be found in the target dataset $\mathcal{D}_t$, but is supposedly smaller.
Hence we have that $|\mathcal{C}_s| < |\mathcal{C}_t|$ and that $\mathcal{C}_s \subset \mathcal{C}_t$.
In a setting of domain adaptation, we further have that $p_t^s \neq p_s$, the target distribution of the known source classes
A metric for measuring how different two domain are is the openness betweeen source and target domain\cite{bendale2015open}, defined as $\displaystyle \mathbb{O} = 1-\frac{\mathcal{C}_s}{\mathcal{C}_t}$.
When $\mathbb{O} > 0$, we're dealing with an OSDA problem.

\subsection{Approach}
\label{sec:apporach}
To tackle the task, we chose to split it in two different steps.
First we have to train the model to separate between the known classes ($\mathcal{C}_s$) and the unknown classes ($\mathcal{C}_{t\setminus s})$ in a reliable enough way.
This is achieved by using a semi-supervised task, by training to model to both recognize a sample class and its correct orientation.
The second step is similar to a classic CSDA problem, where we train the model on a union of both source and target datasets.

\subsection{Rotation Recognition}

Let's denote with $rot({\bf x}, i)$ the rotation of the sample image {\bf x} by $i\times 90$ degrees clockwise.
This is the self-supervised part of our proposed model as rotations $i \in [0, 3]$ can be randomly generated and then predicted.
To avoid situations where the objective orientation of a sample would be a too complicated task, even for a human being, we also feed in input to the model the un-rotated image features.
Alternatively, instead to have the model predict the rotation of a sample for any class, we also try using a different head for each one of the known classes $\mathcal{C_s}$, along with different loss functions.

\subsection{Step I: Sample Separation}

To separate samples we train the model on an enhanced version of $\mathcal{D}_s$, $\tilde{\mathcal{D}_s} = \{({\bf x}_i^s, \tilde{\bf x}_i^s, z)\}$ where $\tilde{\bf x}_i^s$ is the rotated version of ${\bf x}_i^s$ and $z$ is the rotation index.
The network is composed by an extractor $E$ and two heads, $R_1$ and $C_1$ in its standard form.
When using a multi-head rotation predictor, it is composed of $|\mathcal{C}_s|+1$ heads for the rotation task, and an additional one for the classification task.
The features of both the original and rotated image are used to predict the rotation as $\tilde{z}_s = softmax(R_1([E({\bf x}^s), E(\tilde{\bf x}^s)]))$ while classes are predicted solely on un-rotated features $\tilde{\bf y}^s = softmax(C_1(E({\bf x}^s)))$.
The model is training by minimizing an objective function defined as $\mathcal{L}_1 = \mathcal{L}_{C_1} + \mathcal{L}_{R_1}$.
Both heads use a Cross-Entropy loss function:

\begin{equation}
  \mathcal{L}_{C_1} = -\sum_{i\in\mathcal{D}_s} {\bf y}_i^s \log \tilde{\bf y}_i^s
\end{equation}
\label{eq:baseloss_class}

\begin{equation}
  \mathcal{L}_{R_1} = -\alpha_1\sum_{i\in\mathcal{\tilde{D}}_s} {\bf y}_i^s \log \tilde{{\bf z}}_i^s
\end{equation}
\label{eq:baseloss_rot}

We also try using a different loss function for the rotation classifier, which is:

\begin{equation}
  \mathcal{L}_{R_1} = -\alpha_1\sum_{i\in\mathcal{D}_s} {\bf y}_i^s \log \tilde{{\bf z}}_i^s+\lambda_1||{\bf v}_i^s-\gamma({\bf z}_i^s)||_2^2
\end{equation}
\label{eq:center_loss}

In both versions $\alpha_1$ is the weight of the rotation task, $v_i$ is the penultimate layer of the rotation classifier $R_1$ and $\gamma({\bf z}_i)$ is the centroid of the features of class of ${\bf v}_i$.
